---
title: "AFS505 HWK4"
author: "William Ottenheimer"
date: "2022-09-20"
output: html_document
---
##Homework 4
#### Question 1
(a) Create trim.CV function
create a function trim.CV that first removes potential outliers as identified by the criteria above, then computes the coefficient of variation for a vector of numeric data. I may find IQR() and quantile() functions useful

```{r}
### Create a function as described
trim.CV <- function(x){
  ### calculate iqr and quantiles for cutoff values
 iqrange <- IQR(x)
  quant <- quantile(x)
  ### subset vector
  x2<- x > quant[4]+(1.5*iqrange)
   trimupper <- x[!c(x2)]
   x3<- trimupper > (quant[2]) /(1.5*iqrange)
   
  trimmedvect <- trimupper[c(x3)]
  ### calculate and print coefficient of variation
  coefvar <- sd(trimmedvect)/mean(trimmedvect)
  print(coefvar)
    
  return(trimmedvect)
}
### test function ### a vector with 1 and 99 removed should be displayed
testvect <- c(1,99,50,51,52,48)
displaytest <- trim.CV(testvect)
print(displaytest)
```
(b) Compute trim.CV value for sepal width
Recall that part a asked for the argument of trim.CV to be a numeric vector

```{r}
### Import the Iris data and assign to data frame properly
Q1df <- read.csv("C:\\Users\\otten\\Downloads\\iris.data")
names <- read.csv("C:\\Users\\otten\\Downloads\\iris.names")
colnames(Q1df) <- c("Sepal Length","Sepal Width", "Petal Length","Petal Width","Class")
Q1df$Class <- as.factor(Q1df$Class)
### pass sepal width column to trim function
### trim function receives a numeric vector
trimmedsepal <- trim.CV(Q1df$`Sepal Width`)
### display the trimmed vector 
summary(trimmedsepal)
head(trimmedsepal)
```
#### Question 2
use tapply() and aggregate()
(a) use tapply() to compute group standard deviations for petal width, grouping by variety comment on results

```{r}
### import iris data and assign to a dataframe with factors
Q2df <- read.csv("C:\\Users\\otten\\Downloads\\iris.data")
names <- read.csv("C:\\Users\\otten\\Downloads\\iris.names")
colnames(Q2df) <- c("Sepal Length","Sepal Width", "Petal Length","Petal Width","Class")
Q2df$Class <- as.factor(Q2df$Class)
### compute group sd for width by variety
tapply(Q2df$`Petal Width`, INDEX = Q2df$Class , FUN = sd)


```
Brief Comment:
The standard deviations for petal width are .108 for setosa, .197 for versicolor and .274 for virginica. This means setosas are the most consistent and virginicas have the most petal width variation.

(b) use aggregate() with standard specification to compute group medians for all numerical variables in the dataset, grouping by variety and then comment.

```{r}
### use aggregate function to find median of all columns
aggregate(Q2df[,-5],by =list(Q2df$Class) ,FUN = median)
```
Brief comment:
The median values for setosa sepal length, sepal width, petal length, and petal width respectively are 5,3.4,1.5 , and .2
The median values for versicolor sepal length, sepal width, petal length, and petal width respectively are 5.9, 2.8, 4.35, and 1.3
The median values for virginica sepal length, sepal width, petal length, and petal width respectively are 6.5, 3, 5.55, and 2
These median values are relatively close to each other.

#### Question 3
(a) Use the apply function to compute column wise sums of the numerical data in the iris dataset
```{r}
### access the iris dataset
Q3df <- Q2df
apply(Q3df[,-5], MARGIN = 2,FUN = sum)
```
(b) Use a for loop to compute the following ratio for each row
what fraction of observations have ratios > 2.5
```{r}
### create a loop to make the iris ratio variable
dflength <- nrow(Q3df)
lengthvect <- (1:dflength)
Q3df$iris.ratio <- NA
for (i in lengthvect) {

  Q3df$iris.ratio[i] <- (Q3df$`Sepal Length`[i]+Q3df$`Petal Length`[i])/(Q3df$`Sepal Width`[i]+Q3df$`Petal Width`[i])
}
### compute the fraction of observations with ratios > 2.5
Q3df$gr2.5 <- ifelse(Q3df$iris.ratio>2.5,1,0)
fraction1 <- sum(Q3df$gr2.5)/dflength
library(MASS)
fract <- fractions(fraction1)
print(fraction1)
print(fract)

```
(c) repeat 3b using apply instead of a for loop

```{r}
### complete 3b using a apply instead of a for loop
Q3dfc <- Q3df[,-5]
Q3dfc <- Q3dfc[,-5]
Q3dfc<- Q3dfc[,-5]
dflength2 <- nrow(Q3dfc)
### create ratio function
ratfunction <- function(x){
  sepallength <- x[1]
  sepalwidth <- x[2]
  petallength <- x[3]
  petalwidth <- x[4]

    irisratio <- (sepallength+petallength)/(sepalwidth+petalwidth)  

  return(irisratio)
}
Q3dfc$iris.ratio <- apply(Q3dfc, MARGIN = 1,FUN = ratfunction)

### calculate the faction of observations with ratios >2.5

Q3dfc$gr2.5 <- ifelse(Q3df$iris.ratio>2.5,1,0)
fraction2 <- sum(Q3dfc$gr2.5)/dflength2
library(MASS)
fract2 <- fractions(fraction2)
print(fraction2)
print(fract2)


```
The apply function is allegedly meant to save time but it was much easier and faster for me to use the for loop.

#### Question 4
(a) Read ecoli data

```{r}
### read the ecoli data
setwd("C:\\Users\\otten\\Downloads")
Q4df <- read.table("C:\\Users\\otten\\Downloads\\ecoli.data")
### convert appropriate column to factor
Q4df$V9 <- as.factor(Q4df$V9)
summary(Q4df$V9)
```
(b) Extract data types for each column in my ecoli data frame using both the lapply() and sapply() functions and the data.class() function
```{r}
### extract data types using data class function and sapply/lapply
lapply(Q4df,FUN = data.class)
sapply(Q4df, FUN = data.class)

```
Brief Discussion:
Relative to lapply sapply is much more user-friendly and makes the output smoother. It returned a vector as a default and that made the presentation of the output easier to digest.

(c) Compute the sum for each numeric column in your ecoli data using sapply function. 
```{r}
### remove non numeric columns from dataframe
Q4dfc <- Q4df[,-9]
Q4dfc <- Q4dfc[,-1]
### compute sum for each numeric column
sapply(Q4dfc,FUN = sum)
```
Brief Discussion:
The sums have been computed. The apply function necessitates an input for the Margin argument where the sapply function does not. Sapply also takes lists as input which makes it different from apply.

#### Bonus Question

Use mapply() and seq() to generate the following sequences of values with a single command

```{r}
### Use mapply to generate sequence
mapply(seq, c(1,1,1), c(19,16,4.5), c(2,3,.5))
```
