---
title: "AFS505 Hwk1"
author: "William Ottenheimer"
date: "2022-08-31"
output:
  pdf_document: default
  html_document: default
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 1
(a) Engine 1 
Create the vector below.

```{r pressure, echo=FALSE}
#### Create the Vector and populate the data
vect_engine1 <- c(163.00,3.60,10.30,0.58,3.10)
### Name the elements of the vector
names(vect_engine1) <- c("Displacement (cc)","Net Power Out (kW)","Net Torque (Nm)","Oil Capacity (l)","Fuel Tank Capacity (l)" )
#### Display the vector
vect_engine1
### examine the attributes, dimensions, and data types of the vector
attributes(vect_engine1)
typeof(vect_engine1)
str(vect_engine1)


```
Comment: The attribute function reveals the size of the vector, [1:5], it also reports the names of the vector. The type of function discloses the data type for this vector it is double. str shows the data within the vector is numeric, which makes sense because they are doubles for the type. The str shows that the name values of the vector are characters.

(b)Engine 2
As efficiently as possible, create the vector below.

```{r}
### Create the vector with populated data and assigned names upon creation
vect_engine2 <- c("Displacement (cc)"="163","Net Power Out (kW)"="3.6","Net Torque (Nm)"="10.3","Oil Capacity (l)"= "0.58","Fuel Tank Capacity (l)"="3.1")
### Display the Vector
vect_engine2
### examine the attributes, dimensions and data types of the vector
attributes(vect_engine2)
typeof(vect_engine2)
str(vect_engine2)
```
The attributes still list the size and names of the vector. It is still a [1:5]. The data type is character. This is a key difference. To make this code more efficient I assigned the names in the same line of code where I populated the data. The data type is character because the homework prompt required quotations in the vector. Alternatively I could have coerced the previous data to become character values, either way the transformation requires one line of code.


from notes
```{r}
vect_example <- c(1.75, 70)
vect_example

cat("\n")   # this prints a line space

#now assign names for meaning, notice these are character strings
names(vect_example) <- c("height (m)", "weight (kg)")
vect_example

cat("\n")   # this prints a line space
names(vect_example) <- c("height (m)", "weight (kg)")
vect_example
```

Question 2 - Family

Create a vector with names of your family members making clear which name is for your father, wife, husband, son, cat, etc...

```{r}
### create a vector with family member names, name elements so name ownership is clear.
vect_FamNames <- c("Self"="Will","Father"="Bill","Mother"="Denise","Sister"= "Tori","Pet"="Mrs. Rat","Wife"="NA","Husband"="NA","Son"="NA")
vect_FamNames

```
I have created a vector with names of my family members. I think it is clear which name belongs to which family member. There are several family member categories that don't apply to me.

Question 3 - Matrices
a) create the matrix below
```{r}
### create a vector with all the elements of the matrix in a specific order for the default ordering
Q3A_vect <- c(-3.3,8.2,19.7,-1.0,10.5,22.0,1.3,12.8,24.3,3.6,15.1,26.6,5.9,17.4,28.9)
### create a matrix using the data from the vector and assign it the appropriate number of rows/columns
Q3A_mat <- matrix(data = Q3A_vect, nrow = 3, ncol = 5)
### name the columns to match the homework prompt
colnames(Q3A_mat) <- c("AA","BB","CC","DD","EE")
### display the matrix to ensure accuracy
Q3A_mat
### use str to examine the properties as instructed
str(Q3A_mat)

```
After using str I understand my matrix is a [3,5](3 rows, 5 columns) with numeric data. I see the column names are characters and the row names are defaulted.

b) arithemtic with matrix

add 3.3 to the matrix above, and divide by 2.3 show your results and comment.

(Because of the way this is written I plan on adding the value 3.3 to the matrix and then taking the new matrix from that operation and dividing that subsequent matrix by 2.3. I am not sure if the assignment wants to see the original matrix divided by 2.3 but it doesn't sound like that based on the instructions)
```{r}
### Refer to Q3A_mat for the initial matrix operation and add 3.3 to it
Q3B_matsum <- Q3A_mat + 3
### Display matrix for accuracy
Q3B_matsum
### Divide matrix with new totals by 2.3 (division is equivalent to multiplication by the reciprocal)
Q3B_matquotient <- Q3B_matsum * (1/2.3)
### Display matrix for accuracy
Q3B_matquotient

```
My results are shown I believe the arithmetic was done correctly

c) Create the matrix below using prior results to complete the task efficiently
```{r}
### use previous data to make the matrix quickly
Q3C_mat <- matrix(data = Q3A_vect, ncol = 3,nrow = 5)
### Display matrix
Q3C_mat
### Use str to examine the properties as requested
str(Q3C_mat)
```
The matrix is a 5 by 3 matrix (5 rows, 3 columns) and the data type is numeric, specifically doubles. There are no names for the columns.

d) Create the matrix below
Use prior results to complete this task as efficiently as possible

I recognize the new matrix is a truncated version of the previous one so I will coerce the data to integer type.
```{r}
### Recall the previous matrix Q3C_mat and coerce the data type into integer
Q3D_mat <- matrix(data = as.integer(Q3A_vect), ncol = 3,nrow = 5)
### one of the values is different so I will manually change this
Q3D_mat[1,2] <- 21
### Display matrix
Q3D_mat
### use str to examine properties per instructions
str(Q3D_mat)
```
The str command reveals the matrix is a 5 by 3 in terms of dimensions and contains numeric data, in this case integer data.The columns and rows have no names so there is nothing about them elucidated.


I am generally dissatisfied with requesting "efficiency". There is ambiguity behind what it means to be efficient. You could think about time usage or number of lines of code or number of key strokes so the instructions are unclear. I did what was easiest for me and in a way that's efficient.

-Will O
