---
title: "AFS505 S2 HW2"
author: "William Ottenheimer"
date: "2022-10-06"
output:
  pdf_document: default
  html_document: default
---

Use data from 
https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-
data.csv  

### Question 1

a) Generate a log2 plot showing cases per month for European countries with a population over 1million. Highlight six countries with the highest COVID19 case rates for the month of April

```{r}
### install appropriate packages
require(ggplot2)
require(tidyr)
require(magrittr)
require(dplyr)

```
```{r}
#install.packages("gghighlight")
```


```{r}
library(tidyverse)
library(ggplot2)
library(gghighlight)
```


```{r}
### read appropriate data
Q1DF <- read.csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-
data.csv" )
```

```{r}
### I attempted to filter but found an easier way this failed segment of code represents my previous efforts
##Q1DF %>%
###  filter(Q1DF, Q1DF$continent == 'Europe')
###  filter(Q1DF, population >= 1000000)
###  separate(Q1DF, date, into = c('Year','Month','Day'),sep = '-')
###  summary(Q1DF)
```


```{r}
### Filter observations to only include European nations
Q1DFeur <- filter(Q1DF, continent == 'Europe')
### Filter observations to only include populations over 1 million
Q1DFpop <- filter(Q1DFeur, population >= 1000000)
### duplicate date
Q1DFpop$date2 <- Q1DFpop$date
### use date to create month year day variables

Q1DFdate <- separate(Q1DFpop,date, into = c('Year','Month','Day'), sep= '-')
Q1DFdate$location <- as.factor(Q1DFdate$location)
 
### remove na values
Q1DFcomp <- filter(Q1DFdate, !is.na(total_cases_per_million))
Q1DFcomp$monthlycases <- NA
```

We need to sum the cases per million over the years for each month


```{r}
### Export data frame to figure out the top 6 cutoff
write.csv(plotdata,"C:\\Users\\otten\\Desktop\\AFS Repository\\Will-O-AFS505\\hwk2plotdata.csv")
### for April it is 15618050.024 
### for October it is 8886246.167
```


```{r}
### conditional sum to get cases per month for every year

plotdata <- aggregate(Q1DFcomp$total_cases_per_million, by=list(Q1DFcomp$Month, Q1DFcomp$location), FUN = sum)
colnames(plotdata) <- c('Month','Location','Cases')


### create a dummy variable if the country is in the 6 highest cases for april
plotdata$AprilHigh <- ifelse(plotdata$Month == "04" & plotdata$Cases >= 15618050.024,1,0 )

```


now that we have the right data we can create our plot
```{r}
### generate plot
p <- ggplot(data = plotdata, mapping = aes(x = Month, y = Cases, group = Location,colour = Location)) + geom_line() + ylab("Cases per million ") + ggtitle("European COVID19 Monthly Case Rate") + gghighlight(max(AprilHigh) == 1, use_direct_label = FALSE)
p + scale_y_continuous(trans = 'log2')
```

b) Repeat but this time highlight the six countries with the highest covid 19 case rates for the month of Oct

```{r}
### Identify the top six cases for october
plotdata$OctHigh <- ifelse(plotdata$Month == "10" & plotdata$Cases >= 8886246.167,1,0 )

```

```{r}
p2 <- ggplot(data = plotdata, mapping = aes(x = Month, y = Cases, group = Location,colour = Location)) + geom_line() + ylab("Cases per million ") + ggtitle("European COVID19 Monthly Case Rate") + gghighlight(max(OctHigh) == 1,use_direct_label = FALSE)
p2 + scale_y_continuous(trans = 'log2')
```
### Question 2

a) Generate a plot showing deaths per month for European countries with a population over 1 million. Highlight the six countries with the highest COVID 19 death rates for the month of April

```{r}

### my data is the same as last time so I can use Q1DFcomp
### create a variable that is death/cases and then sum that for each month
Q1DFcomp$drate <-Q1DFcomp$total_cases_per_million/Q1DFcomp$total_deaths_per_million

Q1DFcomp <- drop_na(Q1DFcomp, drate)
```




```{r}
### conditional sum to get cases per month for every year

plotdata2 <- aggregate(Q1DFcomp$drate, by=list(Q1DFcomp$Month, Q1DFcomp$location), FUN = sum)
colnames(plotdata2) <- c('Month','Location','DeathRate')


### create a dummy variable if the country is in the 6 highest cases for April
plotdata2$AprilHigh <- ifelse(plotdata2$Month == "04" & plotdata2$DeathRate >= 11661,1,0 )
```



```{r}
### Export data frame to figure out the top 6 cutoff
write.csv(plotdata2,"C:\\Users\\otten\\Desktop\\AFS Repository\\Will-O-AFS505\\hwk2plotdata3.csv")
### for April it is 11661.327712563 
### for October it is 7797.20516837628
```


now that we have the right data we can create our plot
```{r}
### generate plot
p3 <- ggplot(data = plotdata2, mapping = aes(x = Month, y = DeathRate, group = Location,colour = Location)) + geom_line() + ylab("Death Rate (Cases/Deaths) ") + ggtitle("European COVID19 Monthly Death Rate") + gghighlight(max(AprilHigh) == 1, use_direct_label = FALSE)
p3 
```

b) Repeat but this time highlight the six countries with the highest COVID19 death rates for the month of Oct

```{r}
### create a dummy variable if the country is in the 6 highest cases for October
plotdata2$OctHigh <- ifelse(plotdata2$Month == "10" & plotdata2$DeathRate >= 7797.20516837628,1,0 )
```

```{r}
### generate plot
p4 <- ggplot(data = plotdata2, mapping = aes(x = Month, y = DeathRate, group = Location,colour = Location)) + geom_line() + ylab("Death Rate (Cases/Deaths) ") + ggtitle("European COVID19 Monthly Death Rate") + gghighlight(max(OctHigh) == 1, use_direct_label = FALSE)
p4 
```
All the appropriate graphs were created and this assignment is complete


