---
title: "AFS505hwk3"
author: "William Ottenheimer"
date: "2022-09-12"
output:
  pdf_document: default
  html_document: default
---
Question 1: Simple Lists
(a) List 1
Create list 1 shown below

```{r}
### Create the elements of the list (in this case data objects)
object1 <- c("Soils 502")
object2 <- c(1,3,4,5)
object3vect <- c(1:6)
obj3matr <- matrix(data = object3vect, nrow = 2, ncol = 3)
object3 <- t(obj3matr)
### Create the list from the elements
Q1a_list <- list(object1,object2,object3)
### Display and analyze list
Q1a_list
str(Q1a_list)
### Extract the final element of the list as instructed
Q1a_extract1 <- Q1a_list[3]
### Display first extraction
Q1a_extract1
str(Q1a_extract1)
### Extract two elements from the list as instructed
Q1a_extract2 <- Q1a_list[c(T,F,T)]
### Display second extraction
Q1a_extract2
str(Q1a_extract2)
```

(b) list 2
Create the following list
```{r}
### Create the elements of the list (data objects some of which are lists)
object4a <- c("Williams")
object4b <- c("Serena")
object4 <- list(Last = object4a, First = object4b)
object5 <- 36
object6 <- c(wins = 801,losses = 136)
Austopen <- c(2003,2005,2007,2009,2010,2015,2017)
Frenchopen <- c(2002,2013,2015)
Wimbled <- c(2002,2003,2009,2010,2012,2015,2016)
Usopen <- c(1999,2002,2008,2012,2013,2014)
object7 <- list(Australian.Open = Austopen,French.Open = Frenchopen,Wimbledon = Wimbled, US.Open = Usopen)
### Create the list with specific naming conventions
Q1b_list <- list(name = object4,age = object5, record = object6, grand.slam.wins = object7)
### Display the list and structure
Q1b_list
str(Q1b_list)
```
Grand Slam wins aren't structured as a data frame or matrix because they are structured as a list. There are different lengths of atomic vectors so it wouldn't convert easily to data frame or matrix form.

Question 2 cbind,rbind

Create the following vectors and matricies.

```{r}
### Create the following vectors and matrices
Q2vect1 <- c(3,5,6,12)
Q2vect2 <- c(TRUE,FALSE,TRUE,TRUE)
Q2vect3 <- c("A","B","C","D","E")
Q2mat1 <- matrix(seq(2,32,by=3), nrow = 4, ncol = 5)
Q2vect4 <- c(1,0,-1,0,1,-1,0,1)
Q2mat2 <- matrix(data = Q2vect4, nrow = 2,ncol = 4)
### Display all the created data objects
Q2vect1
Q2vect2
Q2vect3
Q2mat1
Q2mat2
```
(a) cbind
create the following outputs using data objects created and a single cbind command
```{r}
### Create the appropriate data object
v2 <- cbind(Q2vect1,Q2vect2)
### Display the data object
colnames(v2) <- NULL
v2
### Create the next appropriate data object
cbind2 <- cbind(Q2vect3,Q2vect2)
### Display the data object
colnames(cbind2) <- NULL
cbind2
### Create the next appropriate data object
cbind3 <- cbind(Q2mat1,Q2vect1)
### Display the data object
colnames(cbind3) <- NULL
cbind3
```
(b) rbind
Create the three outputs using data objects created
```{r}
### Create the appropriate object
rbind1 <- rbind(Q2mat1[,1:4],Q2mat2)
### Display creation
rbind1
### Create the next object
rbind2 <- rbind(Q2vect1,Q2mat2)
### Display creation
rownames(rbind2) <- NULL
rbind2
### Create the next object
rbind3 <- rbind(Q2vect1,Q2vect2,Q2vect3)
### Display creation
rownames(rbind3) <- c("v1","v2","v3")
rbind3
```
Question 3

(a) Read in Iris data
Read in the csv Iris data
Assign column names
make sure columns have the correct data type
use summary(), head(), or view() to avoid dumping a large data set into the console

```{r}
Q3df <- read.csv("C:\\Users\\otten\\Downloads\\iris.data")
names <- read.csv("C:\\Users\\otten\\Downloads\\iris.names")
colnames(Q3df) <- c("Sepal Length","Sepal Width", "Petal Length","Petal Width","Class")
Q3df$Class <- as.factor(Q3df$Class)
str(Q3df)
summary(Q3df)
```
(b) Present Summaries of the following subsets
```{r}
### create subset
Q3subset1 <- Q3df
Q3subset1 <- subset(Q3subset1,Q3subset1$Class =="Iris-versicolor")
### Display subset summary
summary(Q3subset1)
### create next subset
Q3subset2 <- Q3df
Q3subset2 <- subset(Q3subset2,Q3subset2$`Petal Width` > 1.5 | Q3subset2$`Sepal Width` >3.0)
### Display subset summary
summary(Q3subset2)

```
There is a relationship between the width measurements and iris class. Iris versicolor tend to have smaller widths than the other two classes... although there are exceptions to the rule

Question 4
(a) Read in car data
```{r}
### Read car data
Q4df <- read.csv("C:\\Users\\otten\\Downloads\\car.csv", stringsAsFactors = TRUE)
### Display data frame
summary(Q4df)
head(Q4df)
dim(Q4df)

```
(b) remove all observations (rows) with missing values. show dim and summary of data with appropriate observations removed
```{r}
### Remove na values from data frame
Q4bdf <- Q4df
Q4bdf <- na.omit(Q4bdf)
### Display subsetted data frame
summary(Q4bdf)
dim(Q4bdf)
```
(c) remove duplicate observations (rows)
```{r}
### Remove duplicate observations
Q4cdf <- Q4bdf
Q4cdf <- Q4cdf[!duplicated(Q4cdf),]
### Display subsetted data frame
summary(Q4cdf)
dim(Q4cdf)
```

