---
title: "AFS505 Hwk2"
author: "William Ottenheimer"
date: "2022-08-31"
output:
  pdf_document: default
  html_document: default
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 1
(a) Crop Vector
Create the vector below.

```{r}
### Create the vector and populate necessary data
vect_cropa <- c("apples","wheat","soybeans","potatoes","grapes","cherries","almonds","hay","hops")
### Display the vector
vect_cropa
```
(b) Address crop vector

Replace soybeans with oranges

Use indexing to create a subset of the crop vector with only major crops grown in Washington

Use a logical vector to create a subset of the crop vector with only major crops grown in Washington

```{r}
### Replace soybeans with oranges
vect_cropb <- vect_cropa
vect_cropb[3] <- "oranges"
### Display vector to verify change
vect_cropb
### subset only major crops
vect_cropb1 <- vect_cropb[c(1,2,4,5,6,8,9)]
### Display subset vector
vect_cropb1
### create subset using logical vector
vect_cropb2 <- vect_cropb[c(T,T,F,T,T,T,F,T,T)]
### Display subset vector
vect_cropb2


```
(c) Reorder WA crop vector

Save the Washington major crop vector (subset of original vector)
Reorder the Washington major crop vector as shown 
Use order() to sort in alphabetical

```{r}
### Save the vector as a subset (redundant)
vect_cropc1 <- vect_cropb1
### Reorder as shown
### create index of order and pass to order function
index <- c(1,6,4,2,3,7,5)
vect_cropc1 <- vect_cropc1[order(index)]
### Display ordered vector
#vect_cropc1
### conduct alphabetical ordering
vect_cropc2 <- vect_cropb1

cat("\nsort increasing\n")
order(vect_cropc2)
vect_cropc2 <- vect_cropc2[order(vect_cropc2)]
### Display vector alphabetically ordered
vect_cropc2
```

Question 2

(a) address matrix with positive indices
create the matrix below
extract a specific subset
extract another subset

```{r}
### create the matrix
Q2A_vect <- c(1:20)
Q2A_mat <- matrix(data = Q2A_vect, nrow = 5, ncol = 4)
Q2A_mat <- t(Q2A_mat)
### Display the matrix
Q2A_mat
### extract the first subset

Q2A_subs1 <- Q2A_mat[2:4,3:4]
### Display the first subset
Q2A_subs1
### extract the second subset
extractcolumn <-c(1,3,5)
Q2A_subs2 <- Q2A_mat[2,extractcolumn]
### Display the second subset
Q2A_subs2
```
(b) Subset Matrix with negative indices
Extract the following subset
Extract another subset

```{r}
###extract the first subset from the original matrix
Q2B_subs1 <- Q2A_mat[-3,]
### Display the first subset
Q2B_subs1
### Extract the second subset from the original matrix
removecolumns <- c(3,5)
Q2B_subs2 <- Q2A_mat[-2,-removecolumns]
### Display the second subset
Q2B_subs2
```
(c) Replace values in matrix
create a duplicate matrix for each step of these exercises so the original matrix remains unaltered

Step 1 Replace 9,14 and 19 with 88
Step 2 Replace 9,14 and 19 with 23,24,25
Step 3 Remove the value of 8

```{r}
### Create duplicate unaltered matrix
Q2C_unaltMat <- Q2A_mat
### Perform Step 1
Q2C_step1Mat <- Q2C_unaltMat
Q2C_step1Mat[2:4,4] <- 88
### Display altered matrix
Q2C_step1Mat
### Perform Step 2
Q2C_step2Mat <- Q2C_unaltMat
Q2C_step2Mat[2,4] <- 23
Q2C_step2Mat[3,4] <- 24
Q2C_step2Mat[4,4] <- 25
### Display altered matrix
Q2C_step2Mat
### Perform Step 3
Q2C_step3Mat <- Q2C_unaltMat
Q2C_step3Mat[2,3] <- NA
### Display altered matrix
Q2C_step3Mat

```

